Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
incorporating Pat values into Rmd

# Hold on to these for now as I update the rest of the rmd.
demographics <- read.csv("results/tables/denovodemographicsTable.csv")
OverallPValue <- read.csv("results/tables/denovoOverallPTable.csv")
ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
ZScoreCombinedData <- read.csv("results/tables/denovoCombinedData.csv")
ZScoreTurnbaughData <- read.csv("results/tables/denovoTurnbaughData.csv")
AUCRFData <- read.csv("results/tables/denovoAUCRFDataTable.csv")
PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
ModelledPower1PerDiff <- read.csv("results/tables/denovoModelled1PerDiff.csv")
ModelledPower5PerDiff <- read.csv("results/tables/denovoModelled5PerDiff.csv")
ModelledPower10PerDiff <- read.csv("results/tables/denovoModelled10PerDiff.csv")
ModelledPower15PerDiff <- read.csv("results/tables/denovoModelled15PerDiff.csv")
ActualDiff <- read.csv("results/tables/denovoActualDiff.csv")
PowerModelHRR <- read.csv("results/tables/denovoPowerModelHRR.csv")
PowerModelBFRR <- read.csv("results/tables/denovoPowerModelBFRR.csv")
SimulatedN_H <- read.csv("results/tables/denovoSimulatedN_H.csv")
SimulatedN_BF <- read.csv("results/tables/denovoSimulatedN_BF.csv")
RRSimNNeededH <- read.csv("results/tables/denovoRRSimNNeededH.csv")
RRSimNNeededBF <- read.csv("results/tables/denovoRRSimNNeededBF.csv")

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load specific R files to source data from for graphs

@@

```{r Data_Load_Fig4, echo=FALSE, message=FALSE, warning=FALSE}

@@

**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 8 studies that fit our criteria. We applied three analysis pipelines to the data from each study. First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value. We also used the random forest machine learning algorithm to identify community features that would allow us to to predict whether a person was obese. Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F. Finally, we normalized the community diversity and B/F values within each dataset and pooled the data to test for an association with obesity.

**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals. Only one diversity measure (OTU richness (N=1)) was significantly different between non-obese and obese individuals (P<0.05). With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`) and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05). In 7/8 of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies. When we pooled the results across studies, there was no significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(10^test3R$LogFC[1],2), nsmall=2)`, CI = `r format(round(10^ShannonRes$ci.lb, 2), nsmall=2)`-`r format(round(10^ShannonRes$ci.ub, 2), nsmall=2)`, P-value = `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`) and B/F ratio (RR = `r format(round(10^test3BF$LogFC[1],2), nsmall=2)`, CI = `r format(round(10^BFRes$ci.lb, 2), nsmall=2)`-`r format(round(10^BFRes$ci.ub, 2), nsmall=2)`, P-value = `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`). When we normalized the Shannon diversity data and pooled studies there was also no significant difference between non-obese and obese individuals (difference in mean z-score = `r format(round(combZHTtest$estimate[1]-combZHTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZHTtest$p.value, 2), nsmall=2)`). There was also no significant difference in the normalized B/F ratio between non-obese and obese individuals (difference in mean z-score = `r format(round(combZLogBFTtest$estimate[1]-combZLogBFTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`).  Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.

@@

**Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}

@@

```{r Table_2, echo=FALSE, message=FALSE, warning=FALSE}

@@

```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
combinedData <- read.csv("results/tables/ROCcombinedData.csv", stringsAsFactors = FALSE)

@@

```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}

@@

```{r Figure_4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
# Section A
p1 <- ggplot(group.data.test, aes(obese, ZH))
p2 <- p1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
p3 <- p2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
p4 <- p3 + xlab("") + ylab("Z-score Normalized H") + scale_x_discrete(breaks=c("No", "Yes"), labels=c("Non-Obese", "Obese"))
p5 <- p4 + guides(fill=FALSE)
p6 <- p5 + theme(axis.line=element_line(colour="black"),
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.border = element_blank(),
                 panel.background = element_blank())
p7 <- p6 + ggtitle('A') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

@@

```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}

@@

```{r Figure_6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}

>>>>

# Hold on to these for now as I update the rest of the rmd.
#demographics <- read.csv("results/tables/denovodemographicsTable.csv")
#OverallPValue <- read.csv("results/tables/denovoOverallPTable.csv")
#ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
#BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
#ZScoreCombinedData <- read.csv("results/tables/denovoCombinedData.csv")
#ZScoreTurnbaughData <- read.csv("results/tables/denovoTurnbaughData.csv")
#AUCRFData <- read.csv("results/tables/denovoAUCRFDataTable.csv")
#PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
#ModelledPower1PerDiff <- read.csv("results/tables/denovoModelled1PerDiff.csv")
#ModelledPower5PerDiff <- read.csv("results/tables/denovoModelled5PerDiff.csv")
#ModelledPower10PerDiff <- read.csv("results/tables/denovoModelled10PerDiff.csv")
#ModelledPower15PerDiff <- read.csv("results/tables/denovoModelled15PerDiff.csv")
#ActualDiff <- read.csv("results/tables/denovoActualDiff.csv")
#PowerModelHRR <- read.csv("results/tables/denovoPowerModelHRR.csv")
#PowerModelBFRR <- read.csv("results/tables/denovoPowerModelBFRR.csv")
#SimulatedN_H <- read.csv("results/tables/denovoSimulatedN_H.csv")
#SimulatedN_BF <- read.csv("results/tables/denovoSimulatedN_BF.csv")
#RRSimNNeededH <- read.csv("results/tables/denovoRRSimNNeededH.csv")
#RRSimNNeededBF <- read.csv("results/tables/denovoRRSimNNeededBF.csv")

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
# Load specific R files to source data from for graphs

@@

```{r Data_Load_Fig4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}

@@

**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 8 studies that fit our criteria. We applied three analysis pipelines to the data from each study. First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value. We also used the random forest machine learning algorithm to identify community features that would allow us to to predict whether a person was obese based on either genus or OTU level. Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F. Finally, we normalized the community diversity and B/F values within each dataset and pooled the data to test for an association with obesity.

**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals. A total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` diversity measurements  were found to be significantly different between non-obese and obese indivdiuals (P < 0.05) (Shannon Diversity (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannon"), ]))`), observed richness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "sobs"), ]))`), and evenness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannoneven"), ]))`)). With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`) and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05). In 7/8 of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies. When we pooled the results across studies, there was no significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(10^test3R$LogFC[1],2), nsmall=2)`, CI = `r format(round(10^ShannonRes$ci.lb, 2), nsmall=2)`-`r format(round(10^ShannonRes$ci.ub, 2), nsmall=2)`, P-value = `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`) and B/F ratio (RR = `r format(round(10^test3BF$LogFC[1],2), nsmall=2)`, CI = `r format(round(10^BFRes$ci.lb, 2), nsmall=2)`-`r format(round(10^BFRes$ci.ub, 2), nsmall=2)`, P-value = `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`). When we normalized the Shannon diversity data and pooled studies there was also no significant difference between non-obese and obese individuals (difference in mean z-score = `r format(round(combZHTtest$estimate[1]-combZHTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZHTtest$p.value, 2), nsmall=2)`). There was also no significant difference in the normalized B/F ratio between non-obese and obese individuals (difference in mean z-score = `r format(round(combZLogBFTtest$estimate[1]-combZLogBFTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`).  Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.

@@

**Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}

@@

```{r Table_2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}

@@

```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
combinedData <- read.csv("results/tables/ROCcombinedData.csv", stringsAsFactors = FALSE)

@@

```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}

@@

```{r Figure_4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
# Section A
p1 <- ggplot(group.data.test, aes(obese, ZH))
p2 <- p1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
p3 <- p2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
p4 <- p3 + xlab("") + ylab("Z-score Normalized H") + scale_x_discrete(breaks=c("No", "Yes"), labels=c("Non-Obese", "Obese"))
p5 <- p4 + guides(fill=FALSE)
p6 <- p5 + theme(axis.line=element_line(colour="black"),
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.border = element_blank(),
                 panel.background = element_blank())
p7 <- p6 + ggtitle('A') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

@@

```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}

@@

```{r Figure_6, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
